Redundant calibration as a complex optimization problem

### 2017-01-05 paper.tex
worked a bit on the introduction

@@

\begin{abstract}
Traditionally redundant calibration is implemented via a hierarchical algorithmic strategy, i.e. a basic firstcal is performed, followed by LOGCAL and LINCAL.  We first show
that LINCAL is effectively the non-linear least-squares gradient-based Gauss-Newton algorithm. Therefore, an easy way to reduce the number of layers used in performing redundant calibration is to employ the Levenberg-Marquardt algorithm instead (which would improve the convergence properties of stock standard LINCAL). 
We further show that instead of splitting our calibration problem into its phase and amplitude components,
which is the approach taken by LINCAL, we can treat the complex unknown model parameters and their complex conjugates as separate unknowns instead. In the latter framework
we make use of the gradient-based complex Levenberg-Marquardt algorithm (not its real valued counterpart) to determine our fundamental update step. 
Lastly, we compare two previously proposed methods to speedup redundant 
calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating 
direction implicit method outperforms the preconditioned conjugate gradient method, however the preconditioned conjugate gradient method is more robust as it would be easier to adapt to other calibration use cases. 
The reason we exclusively use the complex optimization framework in this paper, instead of the phase-amplitude approach, is that in our opinion the complex framework makes it much easier to derive an analytical tractable update step. More importantly however, when we work in this complex framework it becomes trivial to show that the alternating direction implicit method is in actual fact just a 
special case of the more general complex Levenberg-Marquardt algorithm.
\end{abstract}

@@

\section{Introduction}
The propogation effects (i.e. the ionosphere, antenna gains, etc.) a celestial signal experiences along its propogation path is modelled with the radio interforemetric measurement equation (RIME). %\citep{ME1,RRIME1}.
Calibration is the procedure by which we remove the errors that these propogation effects induce onto our intereferometric data and is normally accomplished using a 
least-squares solver. In the case where we only model the antenna gains, which is the most basic form of the RIME, we can define calibration as finding the antenna gains that minimizes the difference betweem our observed and predicted visibilities. There are two recent developments
in calibration which should be explicitly mentioned, since they form the foundation this paper builds on. The first is the discovery of StEfCal \citep{Salvini2014}, which 
is an alternating direction implicit (ADI) skymodel-based calibration algorithm. It works by assuming the gains are known while solving for their conjugates. Now we do the reverse,
assume the conugates are known and solve for the gains. This procedure results in the linearization of the calibration algorithm and reduces the computational complexity 
of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^3)$, where $N$ denotes the number of antennas in the array. The second development is the idea; to not break the calibration problem 
into its real and imaginary parts and then to solve for the real and imaginary parts of the parameters seperately, but to rather treat the antenna gains and their 
conjugates as seperate parameters to estimate \citep{Smirnov2015}. \citet{smirnov2015} also showed that the StEfCal algorithm itself is merely a special case of this
gain-and-conjugate approach.

Two baselines are redundant if their beaseline difference vector have the same lenght and orientation, which imply that they measure the same $uv$-modes. When we use a redundant array layout we dramatically 
reduce the number of unkowns. Redundant calibration is the procedure by which we leverage the fact that we have 
Redundant calibration differs from normal calibration in that it does not require a skymodel. With redundant calibration we do not only solve for the antenna gains, but also 
for the true observed visibilities. We are able to do so, because we have less unkonws than baseline equations. Since we

[Redundant calibration has been in use for a long time. Noordam used redundant calibration to produce high-dynamic range images of 3C84 with the 
WSRT array.] RW The first implementations of redundant calibration made use of the logarithm function, because when the logarithm is applied to the scalar measurement equation redundant 
calibration is transformed into a linear problem cite Wieringa,camps and Lui. This approach is also generally referred to as LOGCAL.
LOGCAL is reviewed in Appendix ??. The next improvement was to perform a firt-order taylor expansion (which is equivalent to the GN algorithm Kurien 2016) of the measurement equation an then to solve for the
perturbations of the model-parameters. This approach is commonly referred to as LINCAL. LINCAL is derived from first principals in Appendix . Two ways of speeding up redundant calibration has also been 
proposed. The first technique makes use of the alternating direction implicit method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities (wijnholds and marthi). Using the conjugate gradient method to exploit the sparsity inherent in the redundant calibration problem, 
and in doing so speedup the algorithm itself was proposed in . A proper comparison between these two approaches has not been done before.

@@

Layered approach, LOGCAL, LINCAL, Stefan's conference paper, Noordam, other ASTRON paper, newest papers on redundant calibration, marthi paper etc...

@@

%  \item Present a general framework which unifies all the techniques developed so far showing they are all related (LINCAL, non-linear estimator, etc ...). They are all non-linear
%  least-squares techniques, i.e. they either employ Gauss-Newton or the Levenberg-Marquardt algorithms. We have to start with LINCAL showing that it is GN, then we have to motivate 

@@

%  framework we can find analytic expressions for the Jacobian, the Hessian and the Jacobian-residual product (which is not even the case for LINCAL). State that the algorithm is 

@@

The aforementioned is achieved by first rewriting Eq.~\eqref{eq:least_squares_red} to
\begin{equation}
\label{eq:least_squares_lincal}
\min_{\bmath{\varrho}} \|\breve{\br}\| = \min_{\bmath{\varrho}} \|\breve{\bd} - \breve{\bv}(\bmath{\varrho})\|, 
\end{equation}
where $\breve{\bd} = [\bd^T,\conj{\bd}^T]^T$, $\breve{\bv} = [\bv^T,\conj{\bv}^T]$ and $\bmath{\varrho}$ is the phase and amplitude parameter vector which we define in Appendix~\ref{sec:lincal}. 
With the aid of Eq.~\eqref{eq:least_squares_lincal} we can derive the LINCAL algorithm (see Appendix~\ref{sec:lincal}).
The Jacobian associated with Eq.~\eqref{eq:least_squares_lincal} is equal to Eq.~\eqref{eq:Jac_lin}. 
At the hand of Appendix~\ref{sec:lincal} it becomes evident that LINCAL is equivalent to a basic GN update step \citep{Kurien2016}. 
Therefore, a straightforward way of improving the convergence properties of stock standard LINCAL would be to simply add the LM damping factor to the basic LINCAL update step.

@@

\section{LOGCAL}
\label{sec:logcal}
We can rewrite Eq.~\eqref{eq:vis_red} to \citep{Liu2010}
\begin{eqnarray}
\ln |d_{pq}| &=& \ln |g_p| + \ln |g_q| + \ln |y_{\phi_{pq}}| + s_{pq} \label{eq:logcal_amp}\\
\angle d_{pq} &=& \angle g_p - \angle g_q + \angle \phi_{pq} + t_{pq} \label{eq:logcal_phase}
\end{eqnarray}
where
\begin{align}
s_{pq} &= \mathscr{R} \left \{1 + \frac{n_{pq}}{g_p\conj{g_q}y_{\phi_{pq}}} \right \}, & t_{pq} &= \mathscr{I}\left \{1 + \frac{n_{pq}}{g_p\conj{g_q}y_{\phi_{pq}}} \right \}.
\end{align}
Moreover, $\mathscr{R}\{*\}$ and $\mathscr{I}\{*\}$ denote the real and imaginary operators.

@@

\section{LINCAL}
\label{sec:lincal}
In this section we derive LINCAL \citep{Liu2010} from first principles.
%Assume that $\eta_p,\eta_q,\widetilde{\eta}_{\phi_{pq}},\varphi_p,\varphi_q$ and $\widetilde{\varphi}_{\phi_{pq}}$ are real valued variables.
We can rewrite Eq.~\eqref{eq:vis_red_vec} to
\begin{equation}
\label{eq:lincal_vec}
\bd = \bv(\bvarrho) + \bn, 
\end{equation}
where 
\begin{align}
\left [ \bv(\bvarrho) \right ]_{\alpha_{pq}} &= e^{\eta_p - i \varphi_p} e^{\widetilde{\eta}_{\phi_{pq}}- i \widetilde{\varphi}_{\phi_{pq}}} e^{\eta_q + i \varphi_q}\\
&= v_{pq},
\end{align}
and
\begin{equation}
\bvarrho = 
\begin{bmatrix}
\bmath{\eta} \\
\bvarphi \\
\widetilde{\bmath{\eta}} \\
\widetilde{\bvarphi}
\end{bmatrix},\qquad
\begin{aligned}
\bmath{\eta} &= [\eta_1, \eta_2,\cdots \eta_N]^T\\
 \bmath{\varphi} &= [\varphi_1, \varphi_2,\cdots \varphi_N]^T\\
 \widetilde{\bmath{\eta}} &= [\widetilde{\eta}_1, \widetilde{\eta}_2,\cdots \widetilde{\eta}_L]^T\\
 \widetilde{\bmath{\varphi}} &= [\widetilde{\varphi}_1, \widetilde{\varphi}_2,\cdots \widetilde{\varphi}_L]^T
\end{aligned}.
\label{eq:help}
\end{equation}
Note that the vectors $\bmath{\eta},~\bvarphi,~\widetilde{\bmath{\eta}}$ and $\widetilde{\bvarphi}$ are real valued.

>>>>

\newcommand{\aaps}{A\&AS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\nat}{Nature}
\newcommand{\physrep}{Phys. Rep.}

@@

\begin{abstract}
Traditionally redundant calibration is implemented via a hierarchical algorithmic strategy, i.e. a basic \textsc{firstcal} is performed, followed by \textsc{logcal} and \textsc{lincal}.  
\textsc{logcal} requires the use of the logartihm, while \textsc{lincal} makes use of the gradient-based Gauss-Newton algorithm.
We show that instead of splitting our calibration problem into its phase and amplitude components,
which is the approach taken by \textsc{lincal}, we can treat the complex unknown model parameters and their complex conjugates as separate unknowns instead. 
We further propose the use of the Levenberg-Marquardt algorithm, instead of the Gauss-Newton algorithm, to improve the convergence properties of the proposed complex
framework, which in turn reduces the number of layers that redundant calibration pipelines require. We also compare two previously proposed methods to speedup redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating 
direction implicit method outperforms the preconditioned conjugate gradient method, however the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. 
We mainly focus on the complex framework in this paper rather than the amplitude and phase approach as it makes it trivial to relate the alternating direction implicit method 
and the Levenberg-Marquardt algorithm with each other.
\end{abstract}

%that \textsc{lincal} is effectively the non-linear least-squares gradient-based Gauss-Newton algorithm. Therefore, an easy way to reduce the number of layers used in performing redundant calibration is to employ the Levenberg-Marquardt algorithm instead (which would improve the convergence properties of stock standard \textsc{lincal})

@@

\section{Introduction}
The propogation effects (i.e. the ionosphere, antenna gains, etc.) a celestial signal experiences along its propogation path is modelled with the radio interforemetric measurement equation (RIME) \citep{ME1,RRIME1}.
Calibration is the procedure by which we remove the errors that these propogation effects induce onto our intereferometric data and is normally accomplished using a 
least-squares solver. In the case where we only model the antenna gains, which is the most basic form of the RIME, we can define calibration as finding the antenna gains that minimizes the difference betweem our observed and predicted visibilities.

There are two recent developments in skymodel-based calibration which should be explicitly mentioned at this point, since they form the foundation this paper builds on. The first is the discovery of \textsc{StEfCal} \citep{Mitchell:MWA-cal,Salvini2014}, which 
is an alternating direction implicit (ADI) skymodel-based calibration algorithm. It works by assuming the gains are known while solving for their conjugates and then performing the reverse. 
This procedure results in the linearization of the calibration algorithm and reduces the computational complexity of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$, where $N$ denotes the number of antennas in the array. 
assume the conugates are known and solve for the gains. The second development is the idea: to not break the calibration problem 
into its real and imaginary parts and then to solve for the real and imaginary parts of the unknown parameters seperately, but to rather treat the antenna gains and their 
conjugates as seperate parameters to estimate \citep{Smirnov2015}. \citet{Smirnov2015} also showed that the \textsc{StEfCal} algorithm itself is merely a special case of this
gain-and-conjugate approach.

Two baselines are redundant if their beaseline difference vectors have the same lenght and orientation, which imply that they measure the same $uv$-modes. When we use a redundant array layout we dramatically 
reduce the number of unkowns. If our array is redundant enough we reduce the number of unknowns to such an extend that we are able to solve for the unknown 
observed visibilities themselves in addition to the antenna gains. Redundant calibration is the procedure by which we estimate the unknown observed visibilities in addition
to the antenna gains.

Redundant calibration was already in use in the 1980's. \citet{Noordam1982} used redundant calibration and the WSRT (Westerbork Synthesis Radio Telescope) to produce high-dynamic range images of 3C84.
The first implementations of redundant calibration made use of the logarithm function, because when the logarithm is applied to the scalar measurement equation, redundant 
calibration is transformed into a linear problem \citep{Wieringa1992,Camps2003,Liu2010}. This approach is commenly referred to as \textsc{logcal}.
We briefly discuss \textsc{logcal} in Appendix~\ref{sec:logcal}. The next improvement in redundant calibration was to perform a firt-order taylor expansion, which is equivalent to the Gauss-Newton (GN) algorithm \citep{Kurien2016}, of the measurement equation an then to solve for the
perturbations of the model-parameters \citep{Liu2010}. The implementation of  this approach is called \textsc{lincal}. \textsc{lincal} is derived from first principals in Appendix~\ref{sec:lincal}. 
\textsc{logcal} and \textsc{lincal} are both implemented in \textsc{omnical}\footnote{https://github.com/jeffzhen/omnical}, which is a layered redundant calibration pipeline \citep{Zheng2014,Ali2015}.
The solutions of \textsc{logcal} are used as initial input for \textsc{lincal}, since the GN algorithm requires that its starting parameter guesses be in the vacinity 
of the solution.

Two ways of speeding up redundant calibration have also been proposed recently. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities \citep{Wijnholds2012,Marthi2014}. The second is to use the preconditioned conjugate gradthient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem, 
to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before.

@@

Layered approach, LOGCAL, \textsc{lincal}, Stefan's conference paper, Noordam, other ASTRON paper, newest papers on redundant calibration, marthi paper etc...

@@

%  \item Present a general framework which unifies all the techniques developed so far showing they are all related (\textsc{lincal}, non-linear estimator, etc ...). They are all non-linear
%  least-squares techniques, i.e. they either employ Gauss-Newton or the Levenberg-Marquardt algorithms. We have to start with \textsc{lincal} showing that it is GN, then we have to motivate 

@@

%  framework we can find analytic expressions for the Jacobian, the Hessian and the Jacobian-residual product (which is not even the case for \textsc{lincal}). State that the algorithm is 

@@

The aforementioned is achieved by first rewriting Eq.~\eqref{eq:least_squares_red} to
\begin{equation}
\label{eq:least_squares_lincal}
\min_{\bmath{\varrho}} \|\breve{\br}\| = \min_{\bmath{\varrho}} \|\breve{\bd} - \breve{\bv}(\bmath{\varrho})\|, 
\end{equation}
where $\breve{\bd} = [\bd^T,\conj{\bd}^T]^T$, $\breve{\bv} = [\bv^T,\conj{\bv}^T]$ and $\bmath{\varrho}$ is the phase and amplitude parameter vector which we define in Appendix~\ref{sec:lincal}. 
With the aid of Eq.~\eqref{eq:least_squares_lincal} we can derive the \textsc{lincal} algorithm (see Appendix~\ref{sec:lincal}).
The Jacobian associated with Eq.~\eqref{eq:least_squares_lincal} is equal to Eq.~\eqref{eq:Jac_lin}. 
At the hand of Appendix~\ref{sec:lincal} it becomes evident that \textsc{lincal} is equivalent to a basic GN update step \citep{Kurien2016}. 
Therefore, a straightforward way of improving the convergence properties of stock standard \textsc{lincal} would be to simply add the LM damping factor to the basic \textsc{lincal} update step.

@@

\section{\textsc{LOGCAL}}
\label{sec:logcal}
We can rewrite Eq.~\eqref{eq:vis_red} to \citep{Liu2010}
\begin{eqnarray}
\ln |d_{pq}| &=& \ln |g_p| + \ln |g_q| + \ln |y_{\phi_{pq}}| + s_{pq} \label{eq:logcal_amp}\\
\angle d_{pq} &=& \angle g_p - \angle g_q + \angle \phi_{pq} + t_{pq} \label{eq:logcal_phase}
\end{eqnarray}
where
\begin{align}
s_{pq} &= \mathscr{R} \left \{1 + \frac{n_{pq}}{g_p\conj{g_q}y_{\phi_{pq}}} \right \}, & t_{pq} &= \mathscr{I}\left \{1 + \frac{n_{pq}}{g_p\conj{g_q}y_{\phi_{pq}}} \right \}.
\end{align}
Moreover, $\mathscr{R}\{*\}$ and $\mathscr{I}\{*\}$ denote the real and imaginary operators.

@@

\section{\textsc{lincal}}
\label{sec:lincal}
In this section we derive \textsc{lincal} \citep{Liu2010} from first principles.
%Assume that $\eta_p,\eta_q,\widetilde{\eta}_{\phi_{pq}},\varphi_p,\varphi_q$ and $\widetilde{\varphi}_{\phi_{pq}}$ are real valued variables.
We can rewrite Eq.~\eqref{eq:vis_red_vec} to
\begin{equation}
\label{eq:lincal_vec}
\bd = \bv(\bvarrho) + \bn, 
\end{equation}
where 
\begin{align}
\left [ \bv(\bvarrho) \right ]_{\alpha_{pq}} &= e^{\eta_p - i \varphi_p} e^{\widetilde{\eta}_{\phi_{pq}}- i \widetilde{\varphi}_{\phi_{pq}}} e^{\eta_q + i \varphi_q}\\
&= v_{pq},
\end{align}
and
\begin{equation}
\bvarrho = 
\begin{bmatrix}
\bmath{\eta} \\
\bvarphi \\
\widetilde{\bmath{\eta}} \\
\widetilde{\bvarphi}
\end{bmatrix},\qquad
\begin{aligned}
\bmath{\eta} &= [\eta_1, \eta_2,\cdots \eta_N]^T\\
 \bmath{\varphi} &= [\varphi_1, \varphi_2,\cdots \varphi_N]^T\\
 \widetilde{\bmath{\eta}} &= [\widetilde{\eta}_1, \widetilde{\eta}_2,\cdots \widetilde{\eta}_L]^T\\
 \widetilde{\bmath{\varphi}} &= [\widetilde{\varphi}_1, \widetilde{\varphi}_2,\cdots \widetilde{\varphi}_L]^T
\end{aligned}.
\label{eq:help}
\end{equation}
Note that the vectors $\bmath{\eta},~\bvarphi,~\widetilde{\bmath{\eta}}$ and $\widetilde{\bvarphi}$ are real valued.
